### 2025-07-15 paper/quarto_aaai/paper.tex
minor adjustments in text

In all four cases, it is possible to generate valid counterfactuals
(stars) for unsuccessful applicants (blue). They cross the decision
boundary (green) into the target class, but their quality differs. In
panel (a), they are not plausible: they do not comply with the
distribution of the factuals in \(y^+\) to the point where they form a
clearly discernible cluster. In panel (b), they are highly plausible,
meeting the first objective of Def. \ref{def-explainability}. This
difference in outcomes is quantified for the non-linear MLP in the first
two columns of Table~\ref{tbl-main} as the \(\%\)-reduction in
implausibility: it is substantial and statistically significant for
\emph{LS} across both metrics, the distance-based \(IP\) (29\%) and
divergence-based \(IP^{*}\) (55\%).

@@

We find that CT generally leads to substantial and statistically
significant improvements in plausibility: average reductions in \(IP\)
range from around 7\% for \emph{MNIST} to almost 60\% for \emph{Circ};
for the real-world tabular datasets they are around 12\% for both
\emph{CH} and \emph{Cred} and almost 25\% for \emph{GMSC}; for
\emph{Adult} and \emph{OL} we find no significant impact of CT on
\(IP\). Reductions in \(IP^{*}\) are even more substantial and generally
statistically significant, although the average degree of uncertainty is
higher than for \(IP\): average reductions range from around 20\%
(\emph{Moons}) to almost 90\% (\emph{Circ}). The only negative findings
for \emph{OL} and \emph{MNIST} are statistically insignificant and, for
MNIST, do not align with qualitative findings, which are much more
plausible for CT (Figure~\ref{fig-mnist}).

>>>>

In all four cases, it is possible to generate valid counterfactuals
(stars) for unsuccessful applicants (blue). They cross the decision
boundary (green) into the target class, but their quality differs. In
panel (a), they are not plausible: they do not comply with the
distribution of the factuals in \(y^+\) to the point where they form a
clearly discernible cluster. In panel (b), they are highly plausible,
meeting the first objective of Def. \ref{def-explainability}. This
difference in outcomes is quantified for the non-linear MLP in the first
two columns of Table~\ref{tbl-main} as the \(\%\)-reduction in
implausibility: it is substantial and statistically significant for
\emph{LS} across both metrics, the distance-based \(\text{IP}\) (29\%)
and divergence-based \(\text{IP}^*\) (55\%).

@@

We find that CT generally leads to substantial and statistically
significant improvements in plausibility: average reductions in
\(\text{IP}\) range from around 7\% for \emph{MNIST} to almost 60\% for
\emph{Circ}; for the real-world tabular datasets they are around 12\%
for both \emph{CH} and \emph{Cred} and almost 25\% for \emph{GMSC}; for
\emph{Adult} and \emph{OL} we find no significant impact of CT on
\(\text{IP}\). Reductions in \(\text{IP}^*\) are even more substantial
and generally statistically significant, although the average degree of
uncertainty is higher than for \(\text{IP}\): average reductions range
from around 20\% (\emph{Moons}) to almost 90\% (\emph{Circ}). The only
negative findings for \emph{OL} and \emph{MNIST} are statistically
insignificant and, for MNIST, do not align with qualitative findings,
which are much more plausible for CT (Figure~\ref{fig-mnist}).

### 2025-07-15 paper/sections/experiments.qmd
minor adjustments in text

In all four cases, it is possible to generate valid counterfactuals (stars) for unsuccessful applicants (blue). They cross the decision boundary (green) into the target class, but their quality differs. In panel (a), they are not plausible: they do not comply with the distribution of the factuals in $y^+$ to the point where they form a clearly discernible cluster. In panel (b), they are highly plausible, meeting the first objective of Def. \ref{def-explainability}. This difference in outcomes is quantified for the non-linear MLP in the first two columns of @tbl-main as the $\%$-reduction in implausibility: it is substantial and statistically significant for *LS* across both metrics, the distance-based $IP$ (29%) and divergence-based $IP^{*}$ (55%).

@@

We find that CT generally leads to substantial and statistically significant improvements in plausibility: average reductions in $IP$ range from around 7% for *MNIST* to almost 60% for *Circ*; for the real-world tabular datasets they are around 12% for both *CH* and *Cred* and almost 25% for *GMSC*; for *Adult* and *OL* we find no significant impact of CT on $IP$. Reductions in $IP^{*}$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $IP$: average reductions range from around 20% (*Moons*) to almost 90% (*Circ*). The only negative findings for *OL* and *MNIST* are statistically insignificant and, for MNIST, do not align with qualitative findings, which are much more plausible for CT (@fig-mnist).

>>>>

In all four cases, it is possible to generate valid counterfactuals (stars) for unsuccessful applicants (blue). They cross the decision boundary (green) into the target class, but their quality differs. In panel (a), they are not plausible: they do not comply with the distribution of the factuals in $y^+$ to the point where they form a clearly discernible cluster. In panel (b), they are highly plausible, meeting the first objective of Def. \ref{def-explainability}. This difference in outcomes is quantified for the non-linear MLP in the first two columns of @tbl-main as the $\%$-reduction in implausibility: it is substantial and statistically significant for *LS* across both metrics, the distance-based $\text{IP}$ (29%) and divergence-based $\text{IP}^*$ (55%).

@@

We find that CT generally leads to substantial and statistically significant improvements in plausibility: average reductions in $\text{IP}$ range from around 7% for *MNIST* to almost 60% for *Circ*; for the real-world tabular datasets they are around 12% for both *CH* and *Cred* and almost 25% for *GMSC*; for *Adult* and *OL* we find no significant impact of CT on $\text{IP}$. Reductions in $\text{IP}^*$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $\text{IP}$: average reductions range from around 20% (*Moons*) to almost 90% (*Circ*). The only negative findings for *OL* and *MNIST* are statistically insignificant and, for MNIST, do not align with qualitative findings, which are much more plausible for CT (@fig-mnist).
